# Simple mapping in R

Make sure you have read the part about projections and CRS from the introduction page before starting.

We will start our code by first ensuring that our R environment and memory is clean and empty, and then we will install and load R libraries (Section 1) and set up variables in R for file paths or frequently used objects (Section 2).

```{r}
#| label: R-garbage-collection
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

rm(list=ls())
gc()
```


### Section 1: Libraries
I wrote a function that I use in most of my projects to download and load R libraries. Usually I would save the function into it's own R-script and load it using the source() function. However, for the purporse of this tutorial it's best to see the actual function and try to understand what it does. It's generally a good idea to do that for functions that have not been written by yourself in R. While the function I wrote is useful when you work a lot in R, you can also do the same job by hand using multiple lines of code: 
install.packages("package1"), install.packages("package2"), ..., library(package1), library(package2), and so on.

```{r}
#| label: helper-functions
#| eval: false
#| include: true
#| echo: true
#| warning: true
#| message: true

# ------------------------------------------------------------------- #
# 1. Install packages if not already installed and load them
# ------------------------------------------------------------------- #

install_if_missing <-
  function(packages) {
    missing <-
      packages[!packages %in% rownames(installed.packages())]

    # first install the missing libraries
    if (length(missing) > 0) {
      install.packages(
        missing,
        repos = "https://cloud.r-project.org"
      )
    }
    # now load the libraries into our active R session
    x <- suppressPackageStartupMessages(
      lapply(packages, require, character.only = TRUE)
    )
    rm(x)
  }

```

```{r}
#| label: list-libraries
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# list all packages needed for the code
package_list <-
  c(
    "here",
    "dplyr",
    "ggplot2",
    "sf",
    "terra",
    "tidyterra",
    "geodata",
    "skimr",
    "cowplot",
    "tmap"
  )


```

```{r}
#| eval: true
#| echo: false
x <- suppressPackageStartupMessages(lapply(package_list, require, character.only = TRUE))
rm(x)

# delete the list object from our r environment
rm(package_list)
```


```{r}
#| label: install-and-load-libraries
#| eval: false
#| include: true
#| echo: true
#| warning: true
#| message: true

# run the function - may require your input in the console
install_if_missing(package_list)

```


### Section 2: Variables
It's good practice to collect and set up some variables that we can use in our code at the top of each script. In that way, if you want to change something, it's sufficient to do it once at the top of the script, instead of multiple times for each occurrence in the code.  Usually I do it for character strings that I am using multiple times, data paths for input reading/output saving and plotting themes or color vectors. You can save a lot of time during post-processing if you set up your code in a smart way. 


```{r}
#| label: set-up-variables
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

## Settings:
sf_use_s2(TRUE) # for spherical geometry in lat/long (not planar)

## Data paths:
path_proj <- here("01_simple_mapping")
path_geodata <- here("01_simple_mapping/data/geodata/")

## Character strings:
wkt_behrm <- 'PROJCS["World_Behrmann",
    GEOGCS["WGS 84",
        DATUM["WGS_1984",
            SPHEROID["WGS 84",6378137,298.257223563,
                AUTHORITY["EPSG","7030"]],
            AUTHORITY["EPSG","6326"]],
        PRIMEM["Greenwich",0],
        UNIT["Degree",0.0174532925199433]],
    PROJECTION["Cylindrical_Equal_Area"],
    PARAMETER["standard_parallel_1",30],
    PARAMETER["central_meridian",0],
    PARAMETER["false_easting",0],
    PARAMETER["false_northing",0],
    UNIT["metre",1,
        AUTHORITY["EPSG","9001"]],
    AXIS["Easting",EAST],
    AXIS["Northing",NORTH],
    AUTHORITY["ESRI","54017"]]'

wkt_mda <- 'PROJCS["ProjWiz_Custom_Transverse_Cylindrical_Equal_Area",
 GEOGCS["GCS_WGS_1984",
  DATUM["D_WGS_1984",
   SPHEROID["WGS_1984",6378137.0,298.257223563]],
  PRIMEM["Greenwich",0.0],
  UNIT["Degree",0.0174532925199433]],
 PROJECTION["Transverse_Cylindrical_Equal_Area"],
 PARAMETER["False_Easting",0.0],
 PARAMETER["False_Northing",0.0],
 PARAMETER["Central_Meridian",46.8457031],
 PARAMETER["Scale_Factor",1.0],
 PARAMETER["Latitude_Of_Origin",0.0],
 UNIT["Meter",1.0]]'

## Plotting themes:
my_theme <- 
  theme(
    plot.background = element_rect(fill = "#ffffff"),
    panel.background = element_rect(fill = "#ffffff"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank())

```

## 1. Retrieving data for plotting a map

There are many software tools for creating maps in R. It's possible to
create interactive .html maps using R libraries such as `tmap` or
`leaflet`. For printing it's better to keep the maps static
(non-interactive). We will start with that and if you want to make them
interactive, feel free to do it in addition!

`geodata` is a fantastic library that combines the access to different
kinds of geospatial datasets under one hood. You can check out all the
data it has here: <https://rspatial.github.io/geodata/>

In order to use it most efficiently, we need to assign a path on your
computer where the geodata should be downloaded to. Here we will save it
to the sub-folder `data` and create another sub-folder named
`geodata`: results/geodata

```{r}
#| label: set-geodata-path
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true


if (!dir.exists(path_geodata)) {
  dir.create(path_geodata, recursive = TRUE)
}

geodata_path(here(path_geodata))
```


### Downloading data

We will start simple by downloading the administrative borders of a country using the r package geodata. We can use the `geodata::gadm()` function (or simply `gadm()`) to
retrieve administrative border for Madagascar, it's regions, districts
and communes.


The default return class of the `gadm()` function is a `SpatVector`
(spatial vector; the native format of the library `terra`).


```{r}
#| label: get-gadm3-madagascar-SpatVector
#| eval: false
#| include: true
#| echo: true
#| warning: true
#| message: true

# administrative level 3: communes
# (includes level 0:country, level 1:region, level 2:district)
# discrete variable: vector

# download the data to the geodata_path location
geodata::gadm("MDG", level = 3)
```


### Loading data into R 

The file format of the administrative borders is already an object specifically for R (.rds). It's a fast way of storing R objects while conserving all kinds of parameters about the data, like it's class or types of columns. 

```{r}
#| label: load-vectors-madagascar
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# load the data from file (SpatVector)
# read administrative borders:
adm3 <-
  readRDS(here(path_geodata, "gadm", "gadm41_MDG_3_pk.rds")) %>%
  select("NAME_1", "NAME_2", "NAME_3")

adm3

```

### Plotting a static map

```{r}
#| fig.show: hold
#| eval: true
#| include: true
#| echo: true
#| warning: false
#| message: true
#| label: plot-adms-mada

# we have lat/lon data in degrees - thus we will use the spherical geometry settings
sf_use_s2(TRUE)

p1 <-
  adm3 %>%
  group_by(NAME_1) %>%
  summarize() %>%
  ggplot() +
  geom_sf(fill = "lightgrey", col = "white") +
  geom_sf_text(aes(label = NAME_1), size = 3) + # label the communes
  ylab("Latitude") +
  xlab("Longitude") +
  my_theme +
  ggtitle("Level 1 - Regions")

p2 <-
  adm3 %>%
  group_by(NAME_2) %>%
  summarize() %>%
  ggplot() +
  geom_sf(fill = "lightgrey", col = "white") +
  geom_sf_text(aes(label = NAME_2), size = 2) + # label the communes
  ylab("Latitude") +
  xlab("Longitude") +
  my_theme +
  ggtitle("Level 2 - Districts")

p3 <-
  adm3 %>%
  ggplot() +
  geom_sf(fill = "lightgrey", col = "white") +
  geom_sf_text(aes(label = NAME_3), size = 1) + # label the communes
  ylab("Latitude") +
  xlab("Longitude") +
  my_theme +
  ggtitle("Level 3 - Communes")
```

```{r}
#| label: inspect-gadm3-madagascar
#| eval: true
#| include: true
#| echo: true
#| warning: false
#| message: true
#| fig-show: hold
#| fig-width: 12
#| fig-height: 5


# overview of vector data:
adm3
glimpse(adm3)


# combine all three plots into a single panel-plot
par(mfrow = c(1, 1))
plot_grid(p1, p2, p3, labels = c('A', 'B', 'C'), ncol = 3, nrow = 1)
```

### Plotting an interactive map
Make a map with tmap that can be zoomed in and out

```{r}
#| label: plot-tmap-interactively
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true
tmap_mode(mode = "view")


tm_shape(adm3) +
  tm_borders(fill = "NAME_1") +
  tm_text("NAME_3")

```
Let's look at different projections and how they change the shape of Madagascar. 
Below we are comparing WGS84 (the GPS standard in latitude and longitude), equal-area Behrmann projection (global, in Meters), and equal-area projection centered on Madagascar (by projectionwizard.org)

```{r}
#| label: projection-plots
#| eval: true
#| include: true
#| echo: true
#| warning: false
#| message: false
#| fig-show: hold
#| fig-width: 6
#| fig-height: 12
#| layout-ncol: 3

tmap_mode(mode = "plot")
par(mfrow = c(3,1))

# WGS 84 - lat/long
tm_shape(adm3, component.autoscale = FALSE) +
  tm_borders() +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(breaks = c(0, 15, 30), text.size = 1) +
  tm_minimap(height = 3, width = 3)

# Behrman planar projection - meters
tm_shape(adm3) +
  tm_borders() +
  tm_crs(wkt_behrm) +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(breaks = c(0, 15, 30), text.size = 1) +
  tm_minimap(height = 3, width = 3)

# projectionwizard: equal area, centered on MDG
tm_shape(adm3) +
  tm_borders() +
  tm_crs(wkt_mda) +
  tm_compass(position = c("left", "top")) +
  tm_scalebar(breaks = c(0, 15, 30), text.size = 1) +
  tm_minimap(height = 3, width = 3)

```

