# Download environmental data in R

Make sure you have read the part about projections and CRS from the introduction page before starting.

We will start our code by first ensuring that our R environment and memory is clean and empty, and then we will install and load R libraries (Section 1) and set up variables in R for file paths or frequently used objects (Section 2).

```{r}
#| label: R-garbage-collection
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

rm(list=ls())
gc()
```

### Section 1: Libraries
I wrote a function that I use in most of my projects to download and load R libraries. Usually I would save the function into it's own R-script and load it using the source() function. However, for the purporse of this tutorial it's best to see the actual function and try to understand what it does. It's generally a good idea to do that for functions that have not been written by yourself in R. While the function I wrote is useful when you work a lot in R, you can also do the same job by hand using multiple lines of code: 
install.packages("package1"), install.packages("package2"), ..., library(package1), library(package2), and so on.

```{r}
#| label: helper-functions
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# ------------------------------------------------------------------- #
# 1. Install packages if not already installed and load them
# ------------------------------------------------------------------- #

install_if_missing <-
  function(packages) {
    missing <-
      packages[!packages %in% rownames(installed.packages())]

    # first install the missing libraries
    if (length(missing) > 0) {
      install.packages(
        missing,
        repos = "https://cloud.r-project.org"
      )
    }
    # now load the libraries into our active R session
    x <- suppressPackageStartupMessages(
      lapply(packages, require, character.only = TRUE)
    )
    rm(x)
  }

```

```{r}
#| label: list-libraries
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# list all packages needed for the code
package_list <-
  c(
    "here",
    "dplyr",
    "ggplot2",
    "sf",
    "terra",
    "tidyterra",
    "geodata",
    "skimr",
    "cowplot",
    "tmap"
  )


```


```{r}
#| label: install-and-load-libraries
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# run the function - may require your input in the console
install_if_missing(package_list)

# delete the list object from our r environment
rm(package_list)

```

### Section 2: Variables
It's good practice to collect and set up some variables that we can use in our code at the top of each script. In that way, if you want to change something, it's sufficient to do it once at the top of the script, instead of multiple times for each occurrence in the code.  Usually I do it for character strings that I am using multiple times, data paths for input reading/output saving and plotting themes or color vectors. You can save a lot of time during post-processing if you set up your code in a smart way. 


```{r}
#| label: set-up-variables
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

## Data paths:
path_proj <- here("01_simple_mapping")
path_geodata <- here("01_simple_mapping/data/geodata/")

## Character strings:
wkt_behrm <- 'PROJCS["World_Behrmann",
    GEOGCS["WGS 84",
        DATUM["WGS_1984",
            SPHEROID["WGS 84",6378137,298.257223563,
                AUTHORITY["EPSG","7030"]],
            AUTHORITY["EPSG","6326"]],
        PRIMEM["Greenwich",0],
        UNIT["Degree",0.0174532925199433]],
    PROJECTION["Cylindrical_Equal_Area"],
    PARAMETER["standard_parallel_1",30],
    PARAMETER["central_meridian",0],
    PARAMETER["false_easting",0],
    PARAMETER["false_northing",0],
    UNIT["metre",1,
        AUTHORITY["EPSG","9001"]],
    AXIS["Easting",EAST],
    AXIS["Northing",NORTH],
    AUTHORITY["ESRI","54017"]]'

## Plotting themes:
my_theme <- 
  theme(
    plot.background = element_rect(fill = "#ffffff"),
    panel.background = element_rect(fill = "#ffffff"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank())

```

## Downloading environmental data

Let's download some environmental data. 
1) We will download elevation 
2) We will get a couple of soil content variables from isda
african soil database 
3) We will download some climate data from
worldclim (annual temperature, annual precipitation, precipitation
seasonality) 
4) We will download human footprint data.

```{r}
#| label: get-env-madagascar
#| eval: false
#| include: true
#| echo: true
#| warning: true
#| message: true

## 1. elevation (continuos variable - SpatRaster)
geodata::elevation_30s("MDG")

## 2. african soil data (continuos variable - SpatRaster)
geodata::soil_af_isda(
  var = c(
    "oc", # organic carbon (g kg^-1)
    "c.tot", # total carbon (kg^-1)
    "clay", # clay content (%)
    "sand", # sand content (%)
    "silt", # silt content (%)
    "ph.h2o", # pH in H2O
    "bdr", # bed rock depth (cm)
    "texture", # texture class
    "wpg2" # stone content (%)
  ),
  depth = 20,
  path = path_geodata
)

## 3. worldclim: https://www.worldclim.org/data/bioclim.html
worldclim_country(
  var = "bio",
  country = "MDG",
  res = 0.5
)

## 4. human footprint (either 1993 or 2009)
footprint(year = 2009) # It is expressed on a scale of 0 (low) to 50 (high footprint)
```

We will match the resolution and extent of all layers. Then match the
environmental layers to capture Madagascar with terra::crop() and
terra::mask().

```{r}
#| label: load-vectors-madagascar
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# load the data from file (SpatVector)
# read administrative borders:
adm3 <-
  readRDS(
    here(path_geodata, "gadm", "gadm41_MDG_3_pk.rds")
  ) %>%
  select("NAME_1", "NAME_2", "NAME_3")

```

### Cropping and Masking Rasters

```{r}
#| label: load-rasters-madagascar
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true

# load the data from file (SpatRaster)

# read elevation:
elev <-
  rast(
    here(path_geodata, "elevation", "MDG_elv_msk.tif")
  ) %>%
  crop(., adm3) %>%
  mask(., adm3)
names(elev) <- "elevation"

# read soil:
soil <-
  rast(
    list.files(
      here(path_geodata, "soil_af_isda"),
      full.names = TRUE
    )
  ) %>%
  crop(., adm3) %>%
  mask(., adm3)

names(soil) <- c(
  "brd",
  "ctot",
  "clay",
  "oc",
  "ph",
  "sand",
  "silt",
  "texture",
  "stone"
)

# read climate:
bio <-
  rast(
    here(path_geodata, "climate", "wc2.1_country", "MDG_wc2.1_30s_bio.tif")
  ) %>%
  select("wc2.1_30s_bio_1", "wc2.1_30s_bio_12", "wc2.1_30s_bio_15") %>%
  crop(., adm3) %>%
  mask(., adm3)

climate <-
  c(
    bio$wc2.1_30s_bio_1, # annual temp
    bio$wc2.1_30s_bio_15, # precipiation seasonality
    bio$wc2.1_30s_bio_12 # annual prec
  )

names(climate) <- c("tavg", "prec_seas", "prec")

# footprint
footp <-
  rast(
    here(path_geodata, "landuse", "wildareas-v3-2009-human-footprint_geo.tif")
  ) %>%
  crop(., adm3) %>%
  mask(., adm3)
names(footp) <- "footp"
```

### Checks: geometries and data
```{r}
#| label: compare-rasters-madagascar
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true
# Compare rasters regarding extent and resolution
compareGeom(
  elev,
  soil,
  climate,
  footp,
  res = TRUE,
  ext = TRUE,
  stopOnError = TRUE,
  messages = TRUE
)

# save all to a common raster stack:
env_stack <- c(
  elev,
  soil,
  climate,
  footp
)

# Compare CRS
st_crs(adm3) == st_crs(env_stack)

rm(elev, soil, climate, footp, bio)
```

```{r}
# overview of raster data:
env_stack
glimpse(env_stack %>% na.omit())
plot(env_stack)
```


Plot the variables interactively with tmap

```{r}
#| label: plot-tmap-interactively
#| eval: true
#| include: true
#| echo: true
#| warning: true
#| message: true
tmap_mode(mode = "view")

# Temperature
tm_shape(env_stack$tavg) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.or_rd",
      midpoint = NA,
      n = 10
    )
  )

# Precipitation
tm_shape(env_stack$prec) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.rd_bu",
      midpoint = NA,
      n = 10
    )
  )

# Precipitation Seasonality
tm_shape(env_stack$prec_seas) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.rd_bu",
      midpoint = NA,
      n = 10
    )
  )

# Elevation
tm_shape(env_stack$elevation) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.br_bg",
      midpoint = NA,
      n = 10
    )
  )

# Human Footprint
tm_shape(env_stack$footp) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "viridis",
      midpoint = NA,
      n = 5
    )
  )

# Soil (some examples):
tm_shape(env_stack$ctot) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.br_bg",
      midpoint = NA,
      n = 10
    )
  )

tm_shape(env_stack$oc) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.br_bg",
      midpoint = NA,
      n = 10
    )
  )

tm_shape(env_stack$stone) +
  tm_raster(
    col.scale = tm_scale_continuous(
      values = "brewer.br_bg",
      midpoint = NA,
      n = 10
    )
  )
```
